---
title: "categorical data -- SDSS workshop"
output: html_document
date: "2023-05-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Numeric data

```{r}
library(ade4)
data("clementines")
str(clementines)
```

## Categorical data

```{r}
library(fivethirtyeight)
str(food_world_cup[,1:12])
```

## Warnings


> - words are hard to work with!

> - not a lot of options (esp. for 1 dimension): bar plot, Cleveland dot plot

> - data cleaning takes more time

> - main choices: *which* categories to plot, *order* of categories


## Types of data


> - nominal does not have a fixed category order

> - ordinal does have a fixed category order

> - ("real") discrete, small ## of possibilities

> - Not always clearcut: nominal vs. ordinal, ordinal vs. discrete, etc.

> - Sometimes numbers = nominal, not discrete


## Ordinal data


Sort in logical order of the categories (left to right)

```{r}
library(tidyverse)
ggplot(food_world_cup, aes(knowledge)) +
    geom_bar(fill = "cornflowerblue") +
    ggtitle("Knowledge level of respondents") +
    theme_bw(16) +
    theme(panel.grid.major.x = element_blank())
```


## Ordinal data


Sort in logical order of the categories (starting at bottom OR top)

```{r}
ggplot(food_world_cup, aes(knowledge)) +
    geom_bar(fill = "cornflowerblue") +
    coord_flip() +
    ggtitle("Knowledge level of respondents") +
    theme_bw(16) +
    theme(panel.grid.major.x = element_blank())
```

## Nominal data


Sort from highest to lowest count (left to right, or top to bottom)

```{r}
student <- read_csv("student_data.csv")
## See "School Codes and Descriptions" in SSOL help menu

ggplot(student, aes(fct_infreq(School))) +
    geom_bar(fill = "cornflowerblue") +
    ggtitle("Number of Intro Stats Students by School") +
    xlab("") +
    theme_bw(16) +
    theme(panel.grid.major.x = element_blank())
```

## Nominal data


... or top to bottom 
```{r, fig.width=7.5}
student$School <- fct_recode(student$School,
                             `Barnard College`= "BC",
                             `Columbia College` = "CC",
                             `General Studies Post Bac` = "GN",
                             `General Studies` = "GS",
                             `School of Professional Studies` = "SP")


ggplot(student, aes(fct_rev(fct_infreq(School)))) +
  geom_bar(fill = "cornflowerblue") +
  coord_flip() +
  ggtitle("Number of Intro Stats Students by School") +
  xlab("") +
  theme_bw(14) +
  theme(panel.grid.major.y = element_blank())
```

## Discrete data


```{r}
library(vcd)
df <- data.frame(Saxony)
ggplot(df, aes(x = nMales, y = Freq)) +
  geom_col(color = "black", fill = "thistle") +
  ggtitle("19c Saxony: # of males in families with 12 children") +
  xlab("number of male children") +
  theme_bw(14) +
  theme(panel.grid.major.x = element_blank())
```

## Two geoms for bar charts


> - Binned data (has a count column) `geom_col()`

> - Unbinned data (no count column)  `geom_bar()`


## `geom_col()`


> - Requires an **x** and **y**

> - Intended to be used with one **continuous** and one **discrete** variables but other combinations may also work

## Look at the data


```{r}
#| echo: true
df_binned <- data.frame(state = c("CA", "NY", "TX"),
                        count = c(4, 3, 2))
df_binned
```

## Bar chart with binned data


```{r}
ggplot(df_binned, aes(x = state, y = count)) + 
  geom_col()
```

## `geom_bar()`


> - Requires an **x** or **y**


> - Intended to be used with one **discrete** variable


## Look at the data


```{r}
df_unbinned <- data.frame(state = c("NY", "CA", "TX", "NY", "CA", "CA", "TX", "CA", "NY"))
str(df_unbinned)
```

## Bar chart with unbinned data

```{r}
ggplot(df_unbinned, aes(x = state)) +
  geom_bar()
```


# Cleveland dot plot


## Cleveland dot plot


```{r CDP, fig.height=7}
## Cleveland Dot Plot theme
world <- read_csv("countries2012.csv")
africa <- world |>
  filter(CONTINENT == "Africa")
ggplot(africa, aes(x = GDP, y = fct_reorder(COUNTRY, GDP))) +
  geom_point(color = "blue") +
  ggtitle("Africa: GDP per capita, 2012") +
  ylab("") +
  theme_linedraw() ## works well for dotplots
```


## Cleveland dot plot with multiple dots


Sorted by 1997 fatality rate

```{r, fig.height=7}
library(AER)
data("USSeatBelts")
belts <- USSeatBelts |>
  filter(year %in% c(1983, 1997)) |>
  select(state, year, fatalities)
 
## `fct_reorder2` --> double sort: year, then fatalities
ggplot(belts, aes(x = fatalities, y = fct_reorder2(state, year == 1997, fatalities,
                                                   .desc = FALSE), color = year)) +
  geom_point() +
  ggtitle("# of fatalities per million traffic miles") +
  ylab("") +
  theme_linedraw()
```

