---
title: "forcats -- SDSS workshop 2023"
output: html_document
date: May 23, 2023

---
```{r setup, include=FALSE, cache=FALSE}
library(knitr)
options(digits = 3, scipen = 999)
opts_chunk$set(fig.align='center',
               message = FALSE, cache = TRUE,
               warning = FALSE, echo = TRUE)
```

```{r, echo = FALSE}
library(tidyverse)
mycolor <- "#002448"; myfill = "#7192E3"
```


# Character vs factor data

character data: plotted alphabetically

factor data: plotted in order of factor levels

```{r}
df <- tibble(chardata = c("cold", "warm", "hot", "hot", "warm", "warm", "cold", "cold", "cold"), 
    factordata = factor(c("cold", "warm", "hot", "hot", "warm", "warm", "cold", "cold", "cold"), 
                                 levels = c("cold", "warm", "hot")))
```



```{r, echo = FALSE, fig.width = 8, fig.height = 3.5}
library(gridExtra)
g1 <- ggplot(df, aes(chardata)) +
  geom_bar(color = mycolor, fill = myfill) + 
  ggtitle("Default: alphabetical order") +
  xlab("Character variable") +
  theme_grey(12)

g2 <- ggplot(df, aes(factordata)) +
  geom_bar(color = mycolor, fill = myfill) + 
  ggtitle("Default: factor level order") +
  xlab("Factor variable") +
  theme_grey(12)
grid.arrange(g1, g2, nrow = 1)
```



# Summary of useful **forcats** functions

`fct_recode(x, ...)` -- change names of levels 

`fct_inorder(x)` -- set level order of `x` to row order

`fct_relevel(x, ...)` -- manually set the order of levels of `x` 

`fct_reorder(x, y)` -- reorder `x` by `y`

`fct_infreq(x)` -- order the levels of `x` by decreasing frequency

`fct_rev(x)` -- reverse the order of factor levels of `x`

`fct_explicit_na(x)` -- turn NAs into a real factor level

`fct_lump_n(x)` -- lumps all levels except for the `n` most frequent

`fct_lump_min(x)` -- lumps levels that appear fewer than `min` times


# Recoding factor levels: `fct_recode()`

```{r, echo = TRUE}
df <- data.frame(name = factor(c("STAT-UN-1201", "STAT-GR-5702",
                                 "STAT-GR-5293")),
                 enrollment = c(86, 172, 12))
df2 <- df
levels(df2$name) = c("Intro", "EDAV", "IMLV")
df2

# better approach

df$name <- fct_recode(df$name, Intro = "STAT-UN-1201", EDAV = "STAT-GR-5702",
                      IMLV = "STAT-GR-5293")
df
```



# Set factor level order to row order

```{r}
x <- factor(c("Jack", "Queen", "King", "Ace"))
levels(x)
fct_inorder(x)
```


# Using `fct_relevel()` to move levels to the beginning

```{r}
x <- c("A", "B", "C", "move1", "D", "E", "move2", "F")

fct_relevel(x, "move1", "move2")
```

# Using `fct_relevel()` to move levels after an item (by position)

```{r}
x <- c("A", "B", "C", "move1", "D", "E", "move2", "F")

fct_relevel(x, "move1", "move2", after = 4) # move after the fourth item
```

# Using `fct_relevel()` to move levels to the end

```{r}
x <- c("A", "B", "C", "move1", "D", "E", "move2", "F")

fct_relevel(x, "move1", "move2", after = Inf)
```


# Set factor level to the order of another variable

```{r}
levels(iris$Species)
ggplot(iris, aes(x = fct_reorder(Species, Sepal.Width, median), y = Sepal.Width)) +
  geom_boxplot() + 
  coord_flip()
```


# Order factor levels by reverse frequency count

```{r}
x <- factor(c("Jack", "Queen", "King", "Ace", "Queen", "King", "King"))
levels(x)
fct_infreq(x)
```


# Binned data

```{r}
df <- data.frame(quarter = factor(c("Q1", "Q2", "Q3", "Q4")),
                 sales = c(213, 125, 421, 315))
df
levels(df$quarter)
```

# Binned, ordinal data, correct level order

```{r}
ggplot(df, aes(x = quarter, y = sales)) + 
  geom_col(color = mycolor, fill= myfill) +
  theme_grey(16)
```


# Binned, ordinal data, levels out of order

INCORRECT

```{r}
df <- data.frame(temperature = factor(c("cold", "warm", "hot")),
                 count = c(15, 5, 22))

ggplot(df, aes(x = temperature, y = count)) +
  geom_col(color = mycolor, fill = myfill) +
  theme_grey(16)
```



# Binned, ordinal data, levels out of order

INCORRECT

```{r}
Births2015 <- tibble(MotherAge = c("15-19 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", 
                                   "40-44 years", "45-49 years", "50 years and over", "Under 15 years"),
                     Num = c(229.715, 850.509, 1152.311, 1094.693, 527.996, 111.848, 8.171, .754, 2.5))

ggplot(Births2015, aes(MotherAge, Num)) + 
  geom_col(color = mycolor, fill = myfill) + 
  ggtitle("United States Births, 2015", subtitle = "in thousands") +
  scale_y_continuous(breaks = seq(0, 1250, 250)) +
  coord_flip() +
  theme_grey(16)
```



# Binned, nominal

INCORRECT

```{r}
pack1 <- data.frame(
  color = c("blue", "brown", "green", "orange", "red", "yellow"),
  count = c(13, 7, 12, 9, 7, 8)
)

ggplot(pack1, aes(color, count)) + 
  geom_col() +
  theme_grey(16)
```


# Binned, nominal (horizontal bars)

INCORRECT

```{r}
ggplot(pack1, aes(color, count)) + 
  geom_col() + 
  coord_flip() +
  theme_grey(16)
```


# Unbinned, ordinal, correct level order

```{r}
# data available here: https://github.com/jtr13/data
student <- read.csv("student_data.csv") # or use readr::read_csv()
glimpse(student)
levels(student$Level)
levels(factor(student$Level))
```


# Unbinned, ordinal, correct level order

```{r}
ggplot(student, aes(Level)) + 
  geom_bar()  +
  ggtitle("Number of Students by Level") +
  theme_grey(16) +
  theme(panel.grid.major.x = element_blank())
```

# Unbinned, ordinal, levels out of order

INCORRECT

```{r}
df <- tibble(temperature = factor(c("cold", "warm", "hot", "hot", "warm",
                                    "warm", "cold", "cold", "cold")))
df %>% 
  ggplot(aes(temperature)) +
  geom_bar() +
  theme_grey(16)
```


# Unbinned, nominal data

```{r, echo = FALSE}
colors <- c("blue", "orange", "green", "yellow", "red", "brown")
# source of probabilities: https://blogs.sas.com/content/iml/2017/02/20/proportion-of-colors-mandms.html
probs2008 <- c(.24, .2, .16, .14, .13, .13)
df <- data.frame(mmcolor = sample(colors, 100, prob = probs2008, replace = TRUE))
```

```{r}
dim(df)
head(df, 10)
```


# Unbinned, nominal data


INCORRECT

```{r}
ggplot(df, aes(mmcolor)) + 
  geom_bar() +
  theme_grey(16)
```


# Unbinned, nominal data

INCORRECT

```{r}
ggplot(df, aes(mmcolor)) + 
  geom_bar() +
  coord_flip() +
  theme_grey(16)
```


# Dealing with NAs

```{r}
df <- data.frame(temperature = factor(c("cold", "warm", "hot", NA)),
                 count = c(15, 5, 22, 12))

ggplot(df, aes(x = fct_inorder(temperature), y = count)) +
  geom_col() +
  theme_grey(16)
```

# Dealing with NAs

```{r}
df <- data.frame(temperature = factor(c("cold", "warm", "hot", NA)),
                 count = c(15, 5, 22, 12))

ggplot(df, aes(x = fct_inorder(temperature), y = count)) +
  geom_col() +
  coord_flip() +
  ggtitle("(NA bar is too prominent)") +
  theme_grey(16)
```

# Dealing with NAs

```{r}
df <- data.frame(temperature = factor(c("cold", "warm", "hot", NA)),
                 count = c(15, 5, 22, 12))
df

ggplot(df, aes(x = fct_rev(fct_inorder(temperature)), y = count)) +
  geom_col() +
  coord_flip() + 
  ggtitle("(Reversing the factor levels doesn't help.)") +
  theme_grey(16)
```

# Dealing with NAs

```{r}
df <- data.frame(temperature = factor(c("cold", "warm", "hot", NA)),
                 count = c(15, 5, 22, 12))

df %>% 
  mutate(temperature = fct_explicit_na(temperature, "NA") %>% 
           fct_relevel("NA", "hot", "warm", "cold")) %>%
  ggplot(aes(x = temperature, y = count)) +
  geom_col(color = mycolor, fill = myfill) +
  coord_flip() +
  theme_grey(16)
```

# Binning

```{r}
df <- data.frame(x = sample(c("red", "green", "blue"), 100, replace = TRUE))

ggplot(df, aes(fct_infreq(x))) + geom_bar()

df %>% group_by(x) %>% summarize(n = n())

binned_df <- df %>% count(x)

binned_df

ggplot(binned_df, aes(x = x, y = n, label = x)) + 
  geom_col() +
  geom_text(aes(y = n/2), col = "white")
```



# Rebinning

```{r}
df <- as.data.frame(Titanic)
head(df)
```

```{r}
ggplot(df, aes(Class, Freq)) + 
  geom_col() +
  theme_grey(16)
```

# Rebinning

The problem:

```{r}
ggplot(df, aes(Class, Freq)) + 
  geom_col(color = "grey50", fill = "lightblue") +
  theme_grey(16)
```

# Rebinning

Rebin:
```{r}
df %>% 
  group_by(Class) %>% 
  summarize(Freq = sum(Freq)) %>% 
  ggplot(aes(Class, Freq)) + 
  geom_col(color = "grey50", fill = "lightblue") +
  theme_grey(16)
```

# Natural order bias?

```{r}
df %>% 
  group_by(Class) %>% 
  summarize(Freq = sum(Freq)) %>% 
  ggplot(aes(fct_rev(Class), Freq)) + 
  geom_col(color = "grey50", fill = "lightblue") +
  xlab("") +
  theme_grey(16)
```

# Is `Class` ordinal or nominal?

```{r}
df %>% 
  group_by(Class) %>% 
  summarize(Freq = sum(Freq)) %>% 
  ggplot(aes(fct_reorder(Class, Freq, .desc = TRUE), Freq)) + 
  geom_col(color = "grey50", fill = "lightblue") +
  xlab("") +
  theme_grey(16)
```

# Is `Class` ordinal or nominal?

```{r}
df %>% 
  mutate(Class = fct_recode(Class, Third = "3rd", First = "1st", Second = "2nd")) %>% 
  group_by(Class) %>% 
  summarize(Freq = sum(Freq)) %>% 
  ggplot(aes(fct_reorder(Class, Freq, .desc = TRUE), Freq)) + 
  geom_col(color = "grey50", fill = "lightblue") + xlab("") +
  theme_grey(16)
```

