---
title: "geom_histogram -- SDSS 2023 workshop"
output: html_document
date: "2023-05-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

# HISTOGRAMS

## Look at the data

```{r}
str(faithful)
```

## Basic histogram

```{r}
library(ggplot2)
ggplot(faithful, aes(x = eruptions)) +
  geom_histogram()
```

## Add color and fill

```{r}
ggplot(faithful, aes(x = eruptions)) +
  geom_histogram(color = "blue", fill = "cornflowerblue")
```

## Vary the bin boundaries

```{r}
ggplot(faithful, aes(x = eruptions)) +
  geom_histogram(breaks = seq(1.5, 5.5, .25),
                 color = "blue", fill = "cornflowerblue")
```

## Density histogram

```{r}
ggplot(faithful, aes(x = eruptions)) +
  geom_histogram(breaks = seq(1.5, 5.5, .25),
                 color = "blue", fill = "cornflowerblue")
```

## Exercises

### 1. Frequency histograms

Choose one of the following datasets:

`swiss` -- built-in base R: 47 rows, 6 variables

```{r}
str(swiss)
```

`movies` -- `ggplot2movies` package: 58788 rows, 24 variables

```{r}
# uncomment and run once:
# install.packages("ggplot2movies")
library(ggplot2movies)
str(movies)
```

Draw histograms of individual variables. Experiment with changing the bin boundaries with `breaks =` or set the binwidth with `binwidth = `

```{r}
library(ggplot2)
```

### 2. Density histograms

Chooose one variable and create a density histogram with `geom_histogram(aes(y = after_stat(density)))`

```{r}

```

